---
title: "Correlation between Inflation and Unemployment"
author: " Edison Jiang, Vincent Chau, Yamato Lchii"
date: 2025/11/07
output:
  pdf_document:
    number_sections: yes
fontsize: 11pt
geometry: margin=1in
---
# Abstract
In this project, we examine the empirical relationship between the U.S. inflation rate and the unemployment rate to assess the extent to which contemporary data align with the classical Phillips curve. We use annual macroeconomic data series from the Federal Reserve Economic Data (FRED) and the World Bank, including U.S. CPI inflation, 10-year breakeven inflation, and the unemployment rate. We first construct time series and scatter plots for exploratory visualization of the data, followed by calculating correlation coefficients and fitting regression models. In a small sample of recent years, U.S. 10-year breakeven inflation and the unemployment rate exhibit a strong and significant negative correlation (r ≈ −0.87, p ≈ 0.02), consistent with the "inflation–unemployment trade-off" described by the short-term Phillips curve. In contrast, using the longer-term World Bank series of U.S. CPI inflation and unemployment, the correlation is close to zero and statistically insignificant (r ≈ 0.08, p ≈ 0.51), indicating that this relationship is not stable over time. Log–log regressions of unemployment on inflation show statistically significant elasticity, but the economic significance is small, and the model's explanatory power is low (R² ≈ 0.06). Overall, our results suggest that the negative link between inflation and unemployment does appear in some subsamples but is not a robust or permanent feature of U.S. data, highlighting the importance of sample selection and model specification when using the Phillips curve for policy analysis.

## Intro
The purpose of this project is to see how inflation, and other similar measures of economic health, influence employment both globally and within the United States. We intend to show how specific demographics can be affected by inflation, as well as showing patterns in previous recessions. Furthermore, we will run our own analysis to see if we can achieve similar results to the Phillip's Curve.

## Data
This project's data was pulled from the data catalogs of the World Bank Group and the Federal Reserve Bank of St. Louis. Most are simple 2 variable lists of the recorded data and the time it was collected, allowing us to use multiple sets of data. As this data was collected from a reliable source, it needed minimal cleaning aside from some reformatting and removal of extraneous labels.

## Visualizations 
These early visualizations indicate a positive relationship between inflation and unemployment, which is to be expected.

The negative correlation between breakeven inflation is expected. Breakeven inflation is calculated by subtracting the nominal yield (the gross amount of money gained from interest) by real yield (same as nominal, except with inflation accounted for, to find the value of that money). Increasing breakeven inflation would mean that real yield has decreased, meaning that the value of money has decreased, indicating an increase in inflation.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
source("Scripts/00_requirements.R")
source("Scripts/01_data_load.R")
```

```{r, echo=FALSE, warning=FALSE}
source("scripts/11_Breakev_Inf_By_Unem.R")
cor.test(fred_data$Inflation, fred_data$Unemployment, use = "complete.obs")
ggplot(fred_data, aes(Inflation, Unemployment)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "USA: 10Y Breakeven Inflation vs Unemployment (FRED annualized)",
       x = "10Y Breakeven Inflation (%)", y = "Unemployment (%)")

# World Bank
source("scripts/12_CPI_By_Unem.R")
cor.test(wb_data$Inflation, wb_data$Unemployment, use = "complete.obs")
ggplot(wb_data, aes(Inflation, Unemployment)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "USA: CPI Inflation vs Unemployment (World Bank annual)",
       x = "Inflation (%)", y = "Unemployment (%)")
```




```{r, echo = FALSE, warning = FALSE}

## Fred and World Bank unemployment comparison visualization
## This is just to confirm that both sources are consistent with eachother

merged_sets <- inner_join(fred_unem_y, wb_unem_us %>% select(Year, Unemployment_WB = Unemployment), by = "Year")

ggplot(merged_sets, aes(x = Year)) +
  geom_line(aes(y = Unemployment, color = "Fred Unemployment"), linewidth = 3) +
  geom_line(aes(y = Unemployment_WB, color = "World Bank Unemployment")) +
  labs(title = "Fred Vs World Bank Unemployment Rates", x = "Year", y = "Unemployment Rates")

```
```{r,  echo = FALSE, warning = FALSE}

## Here I am obtaining and graphing the log transformations of Fred's unemployment rates and inflation rates 
## The log transformation first will aid us in handling with variance as well as brings out any linear relationships better.
##


fred_inflation <- fred_inflation %>% drop_na()
fred_ <- fred_unemploy %>% drop_na()

log_fred_infl <- log(fred_inflation$T10YIE)
log_fred_unem <- log(fred_unemploy$UNRATE)
log_fred_infl <- log_fred_infl[1:900]
log_fred_unem <- log_fred_unem[1:900]
log_reg_fred <- lm(log_fred_infl ~ log_fred_unem, data = fred_data)
summary(log_reg_fred)

log_fred_data <- fred_data %>% mutate(log_fred_unem = log(Unemployment), log_fred_infl =  log(Inflation))

ggplot(log_fred_data, aes(x = log_fred_unem, y = log_fred_infl))+
  geom_point(color = "cyan")+
  geom_smooth(method = "lm")+
  labs(x = "Log of Unemployment", y = "Log of Inflation", title = "Log Transformation of Unemployment and Inflation")
  


```

# Analysis
The diagnostic plot of the fit values versus residuals shows that the linear setting can be used as a reasonable first approximation, but there are still unexplained fluctuations that cannot be ignored.

We also further examine the temporal stability of the inflation-unemployment relationship. Since our data are observed on an annual basis, we plot time series of inflation and unemployment with subsample regression for earlier and later periods, respectively. Visually, the negative correlation between breakeven inflation and unemployment is more pronounced in certain years, while the slope that fits the straight line flattens in newer periods. In subsample regression, the absolute value of the breakeven inflation rate coefficient is generally larger (more negative) in the early sample and smaller in the later sample, suggesting that the short-term Phillips curve-style "inflation-unemployment trade-off" may weaken over time. Considering the small number of annual observations in each subsample, these time-based results still need to be interpreted with caution and are more suitable as suggestive evidence than final conclusions.

Overall, our analysis shows a significant short-term negative correlation between expected inflation (breakeven inflation) and unemployment rates in the FRED data; In the World Bank data, the link between real CPI inflation and the unemployment rate is much weaker and cannot be statistically significantly differentiated from zero. By combining correlation tests, linear regression models, and simple time-based comparisons, we show that the strength and even direction of the inflation-unemployment relationship depend on how inflation is measured, the control variables added, and the time period examined. This finding is consistent with contemporary macroeconomics discussion of the Phillips curve: the Phillips curve is more like an empirical relationship that moves and flattens over time than a fixed structural "law".



